Multiple Sclerosis Diagnosis from Natural Language Processing of Patient Text Data

### 2022-09-13 paper.tex
Fixed some formatting issues

\indent Our work has the following important improvements over existing literature. To create our dataset, we utilize a unique approach where we combine data from social media with the data from the MSRegister database, a database of clinically diagnosed MS patients from the UK \citep{Ford:12}. Since the MSRegister database has a relatively large number of MS patients, with this approach we were able to create the largest ever dataset for computerized MS diagnosis from free text. Moreover, existing research in free text MS diagnosis is also limited because most research relies on only one type of predictive classifier. Thus, another way we extend prior research is by training and testing multiple classification models on the dataset and comparing their performance to get the best classifier for our use-case. \\
\indent Our contributions are as follows: \emph{(i)} create the largest ever dataset for computerized MS diagnosis from free text by combining the data from the existing MSRegister dataset with newly collected social media data \emph{(ii)} perform data analysis on this newly created dataset \emph{(iii)} build and train binary classifiers on the dataset to detect multiple sclerosis from patient free text.

@@

To build binary classifiers for detecting multiple sclerosis from free text, we need a dataset containing experiences of both people with and people without MS, in text format. In this section, we describe how such a dataset was constructed by combining the data from social media platforms with data from the MSRegister database \citep{Ford:12}. We then discuss the privacy and ethical issues surrounding the construction of our dataset.

@@

\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MSRegister database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms (footnote with list link) from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from 3037 diagnosed users in our dataset. \\
\indent \textbf{Control group.} To get the experiences of users who do not have MS (that is, the control group), we chose social media platforms because, like the MSRegister database, people share their experiences in text format on these platforms. Twitter and Reddit were chosen as the social media platforms because they are popular platforms where users often post in plain text about their general life experiences. Therefore, from these social media platforms, we construct two separate control groups. The explanation for the need for two control groups and the description of how these control groups were constructed are outlined below. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MSRegister dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns (footnote), examples of which can be found in Figure ****. After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

The risks associated with our data collection and data access methods are minimal. All the posts collected from Twitter and Reddit were publicly available through the Twitter and Reddit API respectively. We replaced social media usernames with user identifier codes to anonymise their identity. Although the MSRegister database contains identifiable information, including the gender and date of birth of the patients, this database was accessed through a secure virtual desktop environment, so the identifiable information remains private. For both the patients in the MSRegister database and the social media users, we made no attempt identify the users with other external sources of information. Therefore, this research was conducted in compliance to the GDBR guidelines of data protection \citep{GDBR:16}.

@@

  \hspace{1em}

@@

@@

\begin{table}
\begin{tabular}{p{0.5\linewidth}p{0.4\linewidth}}
\textbf{Dataset} & \textbf{Average number of tokens per comment}\\
Diagnosed (MS) & {29.1} \\
Control One (Generic) & {14.1} \\
Control Two (Diabetes) & {85.3} \\ 
\end{table}

@@

\indent \textbf{K-Nearest Neighbours (KNN).} For the KNN classifier, we set the value k to 1, 2, 3, 5, 8, 10, and 100. The best accuracy was then recorded in \autoref{tab:results}. \\

@@

@@

By employing a novel approach combining social media data with the MSRegister dataset of MS patients in the UK, we were able to create the largest ever dataset for computerized diagnosis for MS. We trained multiple different models on this dataset, and we were able to achieve a maximum accuracy of 94.3\% when classifying experiences of MS patients against generic social media text data and a maximum accuracy of 92.4\% against experiences of diabetes patients. This result is significant because with the classifiers trained in this paper, it may be possible to not only detect multiple sclerosis from the patients’ clinical notes, but also from their social media posts. Resulting early diagnosis of MS allows immediate measures to be taken and the symptoms of MS to be controlled. \\

>>>>

\indent Our work has the following important improvements over existing literature. To create our dataset, we utilize a unique approach where we combine data from social media with the data from the MS Register database, a database of clinically diagnosed MS patients from the UK \citep{Ford:12}. Since the MS Register database has a relatively large number of MS patients, with this approach we were able to create the largest ever dataset for computerized MS diagnosis from free text. Moreover, existing research in free text MS diagnosis is also limited because most research relies on only one type of predictive classifier. Thus, another way we extend prior research is by training and testing multiple classification models on the dataset and comparing their performance to get the best classifier for our use-case. \\
\indent Our contributions are as follows: \emph{(i)} create the largest ever dataset for computerized MS diagnosis from free text by combining the data from the existing MS Register dataset with newly collected social media data \emph{(ii)} perform data analysis on this newly created dataset \emph{(iii)} build and train binary classifiers on the dataset to detect multiple sclerosis from patient free text.

@@

To build binary classifiers for detecting multiple sclerosis from free text, we need a dataset containing experiences of both people with and people without MS, in text format. In this section, we describe how such a dataset was constructed by combining the data from social media platforms with data from the MS Register database \citep{Ford:12}. We then discuss the privacy and ethical issues surrounding the construction of our dataset.

@@

\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MS Register database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms (footnote with list link) from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from 3037 diagnosed users in our dataset. \\
\indent \textbf{Control group.} To get the experiences of users who do not have MS (that is, the control group), we chose social media platforms because, like the MS Register database, people share their experiences in text format on these platforms. Twitter and Reddit were chosen as the social media platforms because they are popular platforms where users often post in plain text about their general life experiences. Therefore, from these social media platforms, we construct two separate control groups. The explanation for the need for two control groups and the description of how these control groups were constructed are outlined below. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MS Register dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns (footnote). After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

The risks associated with our data collection and data access methods are minimal. All the posts collected from Twitter and Reddit were publicly available through the Twitter and Reddit API respectively. We replaced social media usernames with user identifier codes to anonymise their identity. Although the MS Register database contains identifiable information, including the gender and date of birth of the patients, this database was accessed through a secure virtual desktop environment, so the identifiable information remains private. For both the patients in the MS Register database and the social media users, we made no attempt identify the users with other external sources of information. Therefore, this research was conducted in compliance to the GDBR guidelines of data protection \citep{GDBR:16}.

@@

  \hspace{0.5em}

@@

\begin{tabular}{lcc}
\textbf{Dataset} & \textbf{\hspace{1.5cm}Average comment length (in number of tokens)\hspace{1.1cm}} & \hspace{1cm}\\
Diagnosed (MS) & 29.1 & \\
Control One (Generic) & 14.1 & \\
Control Two (Diabetes) & 85.3  & \\ 

@@

@@

\indent \textbf{K-Nearest Neighbours (KNN).} For the KNN classifier, we set the hyperparameter $k$ to 1, 2, 3, 5, 8, 10, and 100. The best accuracy was then recorded in \autoref{tab:results}. \\

@@

\begin{table*}[!ht]

@@

By employing a novel approach combining social media data with the MS Register dataset of MS patients in the UK, we were able to create the largest ever dataset for computerized diagnosis for MS. We trained multiple different models on this dataset, and we were able to achieve a maximum accuracy of 94.3\% when classifying experiences of MS patients against generic social media text data and a maximum accuracy of 92.4\% against experiences of diabetes patients. This result is significant because with the classifiers trained in this paper, it may be possible to not only detect multiple sclerosis from the patients’ clinical notes, but also from their social media posts. Resulting early diagnosis of MS allows immediate measures to be taken and the symptoms of MS to be controlled. \\
